---
title: "Preparing newdata for predictions in perfect prog downscaling experiments"
author: "J. Bedia (Santander Meteorology Group)"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        fig_caption: yes
        toc: yes
        number_section: true
vignette: >
  %\VignetteIndexEntry{Preparing newdata in perfect prog downscaling experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

# Data

```{r}
require(transformeR)
```


## Predictand
```{r}
data("VALUE_Iberia_tp")
y <- VALUE_Iberia_tp
```


# Predictors

In this example, we consider the calibration period 1983-1994. To this aim, we first load the example NCEP reanalysis data from package `transformeR`, that encompasses the period 1983-2002 for boreal winter (DJF). `subsetGrid` is used in order to extract the calibration period, indicating the corresponding argument `years`:

```{r}
data("NCEP_Iberia_hus850", "NCEP_Iberia_psl", "NCEP_Iberia_ta850")
mg <- makeMultiGrid(NCEP_Iberia_hus850, NCEP_Iberia_psl, NCEP_Iberia_ta850)
# Calibration period 1983-1994
x <- subsetGrid(mg, years = 1983:1994)
```



# Newdata

We next consider the same NCEP reanalysis to make the predictions, using a different period (1995-2002) than for the calibration (1983-1994). Note that this is just for illustration purposes, and that newdata will be typically formed by GCM simulations (e.g. seasonal to multidecadal...) instead of the reanalysis. For this reason, `prepare_newdata` is envisaged to transparently deal with multimember datasets

```{r}
# Simulation period 1995:2002
prediction.data <- subsetGrid(mg, years = 1995:2002) #%>% redim(member = TRUE)
```



# Procedure

## Preparing the predictor

In this example, the following predictor configuration will be used:

 * Sea-level pressure (`psl`) and air temperature at 850 mb pressure level (`ta850`) are used as global predictors. Their principal components are used explaining the 90\% of variance of psl and 95\% of `ta850`
 * Furthermore, specific humidity at 850 mb (`hus850`) will be used as local predictor. In this example the 4 nearest points to each predictand location are averaged to this aim.
 
The predictor so configured is constructed with `prepare_predictors`:

```{r}
predictor <- prepare_predictors(x = x,
                                y = y,
                                global.vars = c("psl", "ta850"),
                                PCA = list(v.exp = c(.9, .95)),
                                local.predictors = list(neigh.vars = "hus850",
                                                        n.neighs = 4,
                                                        neigh.fun = list(FUN = "mean")))

```

## Preparing the newdata

### Local scaling of the prediction data

Use `help("localScaling", package = "transformeR")` for further details on how this step operates. 

```{r}
newdata <- localScaling(grid = prediction.data, base = x, ref = prediction.data, time.frame = "monthly")
```

Once locally scaled, the newdata is ready to perform the predictions


# Internal structure of newdata

```{r}
str(newdata)
```



